---
title: "Getting started with sdmApp"
author: "Aboubacar HEMA"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting started with sdmApp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.2,
  fig.align = "center",
  dpi = 96,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

`sdmApp` is an R package containing a Shiny application that lets non-expert
R users model species distribution. It brings a reproducible workflow for
species distribution modeling into a single, user friendly environment.
`sdmApp` takes raster data (in any format supported by the `raster` package)
and species occurrence data (several formats supported) as input, and
provides an interactive graphical user interface (GUI).

This vignette has two parts. The first shows the small set of mapping
helpers that `sdmApp` exports, using the example data bundled with the
package, so that you can reproduce every figure below on your own machine.
The second describes the graphical interface and the data it expects.

The main features of the GUI are:

* uploading data (raster predictors and species occurrence files);
* viewing the correlation between raster predictors;
* using [CENFA](https://github.com/rinnan/CENFA) to select species predictors;
* applying spatial blocking for cross-validation, based on the
  [blockCV](https://CRAN.R-project.org/package=blockCV) package;
* fitting species distribution models with or without a spatial blocking
  strategy;
* exporting results;
* staying reproducible by downloading the underlying R code.

The GUI is built around five main windows, selectable from the navigation bar
at the top of the screen. Some of these windows start empty and fill in once
raster and species occurrence data have been uploaded.

# Installation

`sdmApp` is not currently on CRAN, so install it from GitHub. `CENFA` is
also off CRAN and must be installed first.

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("rinnan/CENFA")
remotes::install_github("Abson-dev/sdmApp", dependencies = TRUE)
```

To use the MaxEnt model you additionally need a Java JDK (>= 8), the `rJava`
package, and the `maxent.jar` file placed in the `java` directory of the
`dismo` package, which you can locate with:

```{r maxent-path, eval = FALSE}
system.file("java", package = "dismo")
```

Everything shown in this vignette works without Java.

```{r library}
library(sdmApp)
```

# Launching the interface

A single call starts the application in your browser:

```{r launch, eval = FALSE}
sdmApp()
```

`sdmApp()` accepts a few arguments. `maxRequestSize` sets the largest
upload accepted, in megabytes. `theme` selects the style sheet, one of
`"IHSN"` (the default), `"yeti"`, `"journal"` or `"flatly"`. Setting
`shiny.server = TRUE` returns the application as an object rather than
launching it, which is what you want when deploying behind shiny-server.

```{r launch-args, eval = FALSE}
# accept uploads up to 200 MB and use the flatly theme
sdmApp(maxRequestSize = 200, theme = "flatly")

# return the app object instead of running it
app <- sdmApp(shiny.server = TRUE)
```

Before starting, `sdmApp()` checks that the modelling engines it relies on
are installed, and stops with a message naming any that are missing. This
avoids a cryptic failure halfway through a session.

# The bundled example data

The package ships a small extract of the Niakhar study area in Senegal.

## Occurrence data

`Niakhar.csv` holds 9258 georeferenced trees scored for three species,
*Faidherbia albida*, *Balanites aegyptiaca* and *Anogeissus leiocarpus*
(Ndao et al., 2019). The same table is also provided as `.xlsx`, `.dta`,
`.sav` and `.sas7bdat` so that you can practise uploading each format
through the GUI.

The file is semicolon separated and uses a comma as decimal mark, so it must
be read with `read.csv2()` rather than `read.csv()`.

```{r read-occ}
occ_file <- system.file("extdata", "Niakhar.csv", package = "sdmApp")
occ <- utils::read.csv2(occ_file)

str(occ)
```

The first two columns are longitude and latitude in WGS 84. The remaining
three are binary presence indicators, one per species.

```{r occ-counts}
colSums(occ[, 3:5])
```

For any spatial work, turn the table into an `sf` object. The bundled
rasters are in geographic coordinates, so no reprojection is needed.

```{r occ-sf}
pa_data <- sf::st_as_sf(occ, coords = c("X_WGS84", "Y_WGS84"), crs = 4326)
pa_data[1:3, ]
```

## Environmental variables

The package bundles 34 raster predictors relating to **bioclimatic
drivers**, **soil properties**, **water productivity**, **vegetation
phenology and productivity** and **watershed topography**. In the original
study the predictors were set on a common projection (WGS 84, UTM zone 28N),
cropped to a common extent and resampled to a common resolution of 250 m.
The extract distributed with the package is aligned on a common grid of
93 by 91 cells in geographic coordinates (EPSG:4326).

```{r list-rasters}
files <- list.files(
  system.file("extdata", package = "sdmApp"),
  pattern = "\\.tif$", full.names = FALSE
)
length(files)
head(files, 10)
```

Load one of them, the actual evapotranspiration and interception layer:

```{r load-raster}
r <- raster::raster(system.file("extdata", "AETI.tif", package = "sdmApp"))
r
```

### Bioclimatic variables

Bioclimatic variables are derived from monthly temperature and rainfall
values to produce more biologically meaningful predictors. They capture
annual trends (mean annual temperature, annual precipitation), seasonality
(annual range in temperature and precipitation) and extreme or limiting
factors (temperature of the coldest and warmest month, precipitation of the
wettest and driest quarters). They were extracted from the WorldClim
database version 2 and are averages for the years 1970 to 2000
(Fick and Hijmans, 2017).

### Soil properties

Seven soil property variables come from the ISRIC World Soil Information
portal, produced by the Africa Soil Information Service (AfSIS) project.
More than 85 thousand samples, from over 28 thousand sampling locations and
covering 1950 to 2012, were used for the spatial predictions
(Hengl et al., 2015).

### Water productivity

Two water productivity variables were retrieved from the FAO WaPOR database.
Net biomass water productivity (NBWP) expresses total biomass production in
relation to the volume of water beneficially consumed through canopy
transpiration over the year, net of soil evaporation. Actual
evapotranspiration and interception (AETI) is the sum of soil evaporation,
canopy transpiration and interception, that is the rainfall intercepted by
leaves and evaporated directly from their surface (FAO, 2020). Both are
agrometeorological variables useful for monitoring how effectively
vegetation converts water into biomass (NBWP) and for analysing the soil-air
interface and plant functioning (AETI).

### Vegetation phenology and productivity

Agroforestry systems are landscapes shaped by interactions between crops,
trees and cropping practices. Two phenological metrics related to the
cropping season were therefore included, derived from 16-day MODIS NDVI time
series (MOD13Q1) using the TIMESAT software (Eklundh and Jonsson, 2011).

### Watershed topography

Topographic variables were derived with the Soil and Water Assessment Tool
(SWAT) from the 30 m NASA SRTM digital elevation model. The SWAT watershed
delineator was used to extract 69 sub-basins within the study area, in a
vector file whose attribute table carries the topographic variable values
(Winchell et al., 2010).

# Working with the exported functions

`sdmApp` exports four functions besides the launcher. They are deliberately
small: each takes a raster (or a set of folds) and returns a `ggplot`
object that you can further customise with the usual `ggplot2` grammar.

## Plotting a continuous raster

`sdmApp_RasterPlot()` samples the layer on a regular grid, so it stays fast
on large rasters, and draws it with a terrain colour ramp.

```{r raster-plot}
sdmApp_RasterPlot(r)
```

Because the result is an ordinary `ggplot` object, you can adjust it:

```{r raster-plot-custom}
sdmApp_RasterPlot(r) +
  ggplot2::labs(subtitle = "Niakhar, Senegal")
```

## Plotting a presence/absence surface

`sdmApp_PA()` expects a binary raster and maps absence in red and presence
in green. Here we threshold the AETI layer to build one.

```{r pa-plot}
pa <- r > 4000
names(pa) <- "AETI above 4000"

sdmApp_PA(pa)
```

## Masking a probability surface

A species distribution model produces a probability of occurrence surface.
`sdmApp_TimesRasters()` multiplies it by a presence/absence surface, so that
only the cells predicted as presence keep their probability. The name of the
first raster is carried over to the result.

```{r times-rasters}
prob <- r / raster::cellStats(r, "max")
names(prob) <- "probability of occurrence"

masked <- sdmApp_TimesRasters(prob, pa)
names(masked)

sdmApp_RasterPlot(masked)
```

Every exported function validates its inputs, so a mistake fails early and
with a readable message rather than deep inside `ggplot2`:

```{r validation, error = TRUE}
sdmApp_TimesRasters(1, 2)
```

## Exploring spatial cross-validation folds

Random splits of spatially structured data give over-optimistic error
estimates (Telford and Birks, 2009; Roberts et al., 2017). `sdmApp` relies
on `blockCV` to build spatially separated folds, and
`sdmApp_fold_Explorer()` draws the training and testing points of a chosen
fold on top of a background raster.

```{r fold-explorer, eval = FALSE}
library(blockCV)

sb <- spatialBlock(
  speciesData = pa_data["Faidherbia.albida"],
  species = "Faidherbia.albida",
  rasterLayer = r,
  theRange = 2000,
  k = 5,
  selection = "random",
  iteration = 100
)

sdmApp_fold_Explorer(sb, r, pa_data["Faidherbia.albida"], num = 1)
```

Note a compatibility caveat. From `blockCV` version 3.0, `spatialBlock()` is
deprecated in favour of `cv_spatial()`, and the object returned by
`cv_spatial()` is no longer of class `SpatialBlock`.
`sdmApp_fold_Explorer()` currently tests for that class, so it works with the
deprecated `spatialBlock()` but not yet with `cv_spatial()`. Support for the
new class is planned.

# A typical session in the GUI

The navigation bar exposes five tabs, which are meant to be visited in order.

1. **Help/About** gives a short description of the application and of the
   modelling options.
2. **Data Upload** takes the raster predictors and the species occurrence
   file. Occurrence data can be supplied as CSV, Excel, Stata, SPSS or SAS.
   Nothing else in the interface is populated until this step succeeds.
3. **Spatial Analysis** covers predictor screening. You can inspect the
   correlation matrix between predictors, run an ecological niche factor
   analysis with `CENFA` to rank predictors by their contribution to
   marginality and specialisation, estimate the range of spatial
   autocorrelation, and build spatial blocks for cross-validation.
4. **Modeling** fits the models. Bioclim, Domain, Mahalanobis, GLM, MaxEnt,
   random forest and support vector machines are available, with or without a
   spatial blocking strategy, and results can be exported as rasters and
   tables.
5. **R-Code** returns the R code corresponding to the session, so the
   analysis can be rerun, version controlled and shared.

A reasonable workflow is therefore: upload, drop strongly correlated
predictors, select predictors with `CENFA`, choose a block size informed by
the spatial autocorrelation range, generate folds, fit one or more models,
compare them, then export both the maps and the R code.

# Session information

```{r session-info}
sessionInfo()
```

# References

Bahn, V., McGill, B.J. (2012). Testing the predictive performance of
distribution models. *Oikos*, 122(3), 321-331.

Eklundh, L., Jonsson, P. (2011). *TIMESAT 3.0 Software Manual*. Lund
University, Sweden.

FAO (2020). *WaPOR database methodology*. Food and Agriculture Organization
of the United Nations, Rome.

Fick, S.E., Hijmans, R.J. (2017). WorldClim 2: new 1 km spatial resolution
climate surfaces for global land areas. *International Journal of
Climatology*, 37(12), 4302-4315.

Hastie, T., Tibshirani, R., Friedman, J. (2009). *The Elements of
Statistical Learning: Data Mining, Inference, and Prediction*, 2nd edition.
Springer, New York.

Hengl, T., Heuvelink, G.B.M., Kempen, B., et al. (2015). Mapping soil
properties of Africa at 250 m resolution: random forests significantly
improve current predictions. *PLoS ONE*, 10(6), e0125814.

Hiemstra, P.H., Pebesma, E.J., Twenhofel, C.J., Heuvelink, G.B. (2009).
Real-time automatic interpolation of ambient gamma dose rates from the Dutch
radioactivity monitoring network. *Computers and Geosciences*, 35(8),
1711-1721.

Ndao, B., Leroux, L., Diouf, A.A., Soti, V., Sambou, B. (2019). A remote
sensing based approach for optimizing sampling strategies in crop monitoring
and crop yield estimation studies. In: Wade, S. (Ed.), *Earth Observations
and Geospatial Science in Service of Sustainable Development Goals*.
Springer, pp. 25-36. [doi:10.1007/978-3-030-16016-6_3](https://doi.org/10.1007/978-3-030-16016-6_3)

O’Sullivan, D., Unwin, D.J. (2010). *Geographic Information Analysis*,
2nd edition. John Wiley and Sons.

Phillips, S.J., Anderson, R.P., Dudik, M., Schapire, R.E., Blair, M.E.
(2017). Opening the black box: an open-source release of Maxent.
*Ecography*, 40(7), 887-893.

Rinnan, D.S., Lawler, J. (2019). Climate-niche factor analysis: a spatial
approach to quantifying species vulnerability to climate change.
*Ecography*, 42, 1494-1503. [doi:10.1111/ecog.03937](https://doi.org/10.1111/ecog.03937)

Roberts, D.R., Bahn, V., Ciuti, S., et al. (2017). Cross-validation
strategies for data with temporal, spatial, hierarchical, or phylogenetic
structure. *Ecography*, 40(8), 913-929.

Telford, R.J., Birks, H.J.B. (2009). Evaluation of transfer functions in
spatially structured environments. *Quaternary Science Reviews*, 28(13),
1309-1316.

Valavi, R., Elith, J., Lahoz-Monfort, J.J., Guillera-Arroita, G. (2019).
blockCV: An R package for generating spatially or environmentally separated
folds for k-fold cross-validation of species distribution models.
*Methods in Ecology and Evolution*, 10, 225-232.
[doi:10.1111/2041-210X.13107](https://doi.org/10.1111/2041-210X.13107)

Winchell, M., Srinivasan, R., Di Luzio, M., Arnold, J. (2010).
*ArcSWAT Interface for SWAT2009: User’s Guide*. Texas Agricultural
Experiment Station, Texas.
